Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-11-08 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update Heckman and imputation flow

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\
Data Analytics and Methods Task Force,\\
European Medicines Agency,\\
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}. Our scope is only simple
multilevel models, to show how imputation can yield less biased
estimates from incomplete clustered data. More complex models can be
accomodated, but are outside the scope of this paper.
}

@@

@@

\caption{\label{tab:clus}Concepts in multilevel methods}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
Concept & Details\\
\midrule
Sample unit & Units of the population from which measurements are taken in a sample.\\
Hierarchical levels & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.\\
Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.\\
Mixed effect & Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).\\
\addlinespace
ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
Stratified intercept & \\
\bottomrule
\end{tabular}
\end{table}

@@

Table 2: Concepts in missing data methods

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng94}. \\
& \\
\end{longtable}

@@

@@

In this example, we demonstrate a multilevel imputation of an intercept
and slope random model with a continuous response. We use the obesity
dataset included in the \texttt{micemd}@ package, a synthetic dataset
that emulates an electronic survey in which individuals are asked to
provide information about their weight and consumption habits in
different countries. To easy the explanations we simulate data for 5
clusters and we reduce the dataset to the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{gender} Gender (0=male, 1=female),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{height} Height in meters,
\item
  \texttt{weight} Weight in kilograms,
\item
  `time`` Response time in minutes (inclusion-restriction variable).
\item
  `FamOb`` Family obesity history (yes or not)
\end{itemize}

@@

In the primary analysis model, BMI serves as the dependent variable,
with Age, Gender, and FamOb as predictors. Researchers, based on
observable data, are considering the inclusion of random intercept
effects, as well as introducing a random slope for the Age variable.

The main model is given by:
\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
We start by loading the data:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> require(patchwork)
R>   myplots <- lapply(1:5, function(i) {
+          ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
+            ggtitle(paste0("Cluster", i))+ 
+       theme(legend.position = ifelse(i==1,"bottom","none"))
+       })
R>   myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1} \end{center}

We observe that the missing pattern is quite similar across the
clusters. However, regarding the weight variable, we notice that is
systematically missing in cluster 3. In order to evaluate if it is
required a imputation method for 1-level or 2-level we assess the
Intraclass Correlation Coefficient (ICC) for the outcome variable, we
use the ``performance'' package:

\begin{CodeChunk}
\begin{CodeInput}
R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
R> performance::icc(Nulmodel)
\end{CodeInput}
\begin{CodeOutput}
# Intraclass Correlation Coefficient

    Adjusted ICC: 0.342
  Unadjusted ICC: 0.342
\end{CodeOutput}
\end{CodeChunk}

Since the ICC is above 0.1 and as the main analyis will be consider a
mixed model, we decide to use two-level (2l) imputation methods. In this
imputation process, we include all predictor variables in the main model
and since BMI is a composite variable. We also incorporate weight and
height variables.

To determine the best imputation method for these variables, we can use
the \texttt{find.defaultMethod}function provided in the \texttt{micemd}
package, which suggests an appropriate method for MAR variables based on
the type of variable, number of observations in the cluster, and number
of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
and ``2l.2stage.norm'' for the age variable. However, after inspecting
the age density plot, we consider modifying its method to
``2l.2stage.pmm.'' For the BMI variable, we employ deterministic
imputation.

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
R> meth_mar["Age"]<-"2l.2stage.pmm" 
\end{CodeInput}
\end{CodeChunk}

For these imputation models, it is necessary to specify the prediction
matrix, with the cluster variable labeled as -2 and the predictor
variable labeled as 2, encompassing all variables. We need to suprime
the variable Time as this variable is not especified in the main model
and modify also in the prediction matrix the relationship between BMI,
weight and height to avoid circular predictions. Then we proced to run
the imputation model.

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
R> pred_mar[,"Cluster"]<- -2 # clustering variable
R> pred_mar[,"Time"]<-0
R> pred_mar[pred_mar==1]<-2
R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
R> ggmice::plot_pred(pred_mar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_pred-1} \end{center}

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  2.409  69.703  82.925  82.476  95.310 158.420 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-2} \end{center}

We are also contemplating the utilization of the ``pmm'' option, given
that the imputed values' weights are lower than those of the observable
values.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar_pmm,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  27.50   71.02   83.82   83.76   95.55  135.98 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mar_pmm)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-2} \end{center}

After confirming convergence, we proceed to save the results for future
use. Users are considering the possibility that the weight variable may
not have been selected randomly. It's likely that an omitted variable,
like self-esteem, could influence this selection. For instance,
individuals with lower self-esteem might have higher weight values,
impacting their willingness to provide honest information due to
embarrassment.

To address this situation, two approaches have been proposed for dealing
with Missing Not at Random (MNAR) data: pattern-mixed models and
selection models. Within pattern-mixed models, methods like the delta
method and more advanced ones like NARFS have been suggested. In
contrast, the selection model approach includes methods such as the
Heckman model, which can be particularly useful in this case. Several
methods, including Galimar2017, Hammon2021, and the recent Munoz method,
designed for two-level data, allowing for variations in intercepts and
exposure effects (random intercept and slope).

To apply the heckman method, the weight variable should be specified as
``2l.2stage.heckman'' found in the micemd package. Additionally, the
prediction matrix needs modification because this method involves
specifying two equations: one for the outcome, describing the incomplete
variable in terms of partially observed predictors (in this case, all
variables from the main model), and the other for the selection model,
explaining the probability of being observed based (R) on certain
variables.

Here for the outcome equation we consider is the same imputation model
that we used for the MAR case (main model).
\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
Regarding the selection equation, we include the same predictors as
those in the primary model, as well as a time variable. Here the time
variable serves as a restriction exclusion variable specifically
explaining the probability of being observed but not affecting the
incomplete value (Weight). In this context, we assume that the time a
user spends completing the survey serves as a proxy for the barriers
they may encounter in survey completion, such as familiarity with the
survey content or internet speed. These factors may lead the user to
skip specific questions or even the entire survey. Also we assume the
time is not have any influence on the weight assigned to the subject.
\[R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
These two equations are jointly estimated under the assumption that the
error terms are interconnected with a bivariate normal distribution. For
a more comprehensive understanding of the model and the exclusion
restriction, you can refer to (Munoz).

To integrate information from both equations, we must adjust the
prediction matrix. The cluster variable remains specified as before
(-2). In this imputation method, all the variables present in both the
selection and outcome equations are included with a random effect.
However, it's essential to distinguish which of these variables appear
in each equation.

In this framework, when a variable is shared between both equations, it
is denoted as (2). Predictors exclusive to the outcome equation are
indicated as (-4), while those exclusive to the selection equation are
labeled as (-3). Consequently, the only alteration needed in the
predictor matrix pertains to the variable ``Time.''

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mnar <- pred_mar
R> pred_mnar["Weight","Time"]<- -3
R> ggmice::plot_pred(pred_mnar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_pred-1} \end{center}

We also need to modify the method of the weight variable

@@

Then we proceed to run the imputation model as before Impute the
missingness:

After executing these imputation procedures, it is essential to assess
convergence and the coherence of the imputed values. Upon examining the
weight variable, we noticed that the imputed range falls outside the
realm of plausible values (as weight should be positive). Consequently,
we are contemplating a different approach, specifically, the use of
``pmm,'' which involves imputing values from donor observations. This
approach ensures that the imputed values remain within the range of
observable values.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  10.26   73.41   86.94   88.27  101.73  186.47 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mnar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-2} \end{center}

We then run the imputation model but this time using the option of pmm,
to assure that weight values are in the range of the observable data,
this can be implemented by seeting the pmm parameter as true.

>>>>

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Munoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\
Data Analytics and Methods Task Force,\\
European Medicines Agency,\\
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Munoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

\Abstract{
Incomplete multilevel data requires careful consideration of the missing
data problem and analysis strategy. In this tutorial, we focus on a
popular strategy for accommodating missingness in multilevel data:
replacing the missing data with one or more plausible values, i.e.,
imputation.Imputation separates the missing data problem from the main
analysis and the completed data can be analyzed as if it has been fully
observed.This tutorial illustrates the imputation of incomplete
multilevel data with the statistical programming language R. We aim to
show how imputation can yield less biased estimates from incomplete
clustered data. We provide practical guidelines and code snippets for
different missing data situations, including non-ignorable missingness
mechanisms. For brevity, we focus on multilevel imputation using chained
equations with the R mice package and its adjacent packages.
}

@@

\newpage

@@

\caption{\label{tab:clus0}Concepts in multilevel methods}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
Concept & Details\\
\midrule
Sample units & Units of the population from which measurements are taken in a sample, e.g., students.\\
Cluster & Variable that specify the cluster or agruppation, e.g., Classroom\\
Hierarchical data & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
Level-1 & Variable that varies within a cluster, eg. Test score\\
Level-2 & Variable that does not vary within a cluster but between, e.g. teacher experience.\\
\addlinespace
Hierarchical model & Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)\\
Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)\\
Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. test score tendency during academic year between students due to no controlled factors such as  genetic,family history\\
Mixed effect & Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).\\
ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
\addlinespace
Stratified intercept & \\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:mechanism}Concepts in missing data methods}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
Concept & Details\\
\midrule
MCAR & Missing Completely At Random, where the probability to be missing is equal across all data entries.\\
MAR & Missing At Random, where the probability to be missing depends on observed information\\
MNAR & Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]\\
\bottomrule
\end{tabular}
\end{table}

@@

\hypertarget{imputation-workflow}{%
\subsection{Imputation workflow}\label{imputation-workflow}}

Below we provide a imputation workflow that can be used in general to
impute cluster data.

\hypertarget{main-analysis}{%
\subsubsection{Main analysis}\label{main-analysis}}

When imputing clustered data, our first consideration should be the
research question and the type of analysis that researchers intend to
perform, assuming that non-incomplete values were present. This initial
assessment provides valuable insights into the data's structure (see the
level table), relationships among variables, and the causes of
missingness.

When multiple imputation is used to deal with missing data, as the
imputation and analysis process is performed separately, it is necessary
that imputation model being congenial with the main analysis model
(Meng, 1994), e.g.~if the main model accounts for the hierarchical
structure also imputation model should do it (Audigier, 2021).

\hypertarget{exploration-of-available-data}{%
\subsubsection{Exploration of available
data}\label{exploration-of-available-data}}

Then it is required to explore available information across clusters.
This can involve evaluating the Intra-class Correlation Coefficient
(ICC) to determine if a one-level imputation method is sufficient,
particularly if there are no substantial differences between clusters.
Additionally, it can be done by visualizing the data using available
package as \textbf{ggmice}, for instance examining density plots can
help us understand the variable type (normal, Poisson, categorical) and
its range of plausible values. Exploring relationships between
covariates through bivariate plots can be valuable in identifying
potential interactions or non-linearities.

In the context of missingness, we should assess patterns of systematic
and sporadic missing values and explore the relationships between the
missingness of an incomplete variable and observable data. This
evaluation allows us to consider the plausibility of Missing at Random
(MAR) mechanisms over Missing Completely at Random (MCAR) methods and
helps in selecting prediction variables to be included in the imputation
model (considering overflux).

\hypertarget{selecting-plausible-imputation-methods}{%
\subsubsection{Selecting plausible imputation
methods}\label{selecting-plausible-imputation-methods}}

Before to implement multiple imputation it is requiered to evaluate if
simpler imputation methods are suffice: - For example, certain
incomplete variables may not necessitate stochastic imputation methods
like MICE. Instead, they can be effectively addressed through deductive
imputation, where incomplete values can be inferred from logic and
deterministic relationships between variables. This approach is
particularly useful for variables that are functions of others, e.g.~a
person's BMI from their weight and height, determining values for
one-level variables from two-level ones, e.g.~in the context of
Individual Participant Data (IPD), incomplete information can be
extracted from metadata, such as deducing incomplete data about abortion
in a country where abortion is illegal, or engaging in cross-temporal or
protocol's deduction, like imputing missing test values for deceased
patients.

\begin{itemize}
\item
  Under monotonic pattern, the use of regression models
\item
  CC when the proportion of missing is minimal \textless5\%.
\item
  Analysis that are robust to missing data, for instance FIML models
  such as mixed model do not require imputation models when only the
  outcome variable is missing.An alternative approach to missing data is
  to use Full Information Maximum Likelihood (FIML). This method does
  not require the imputation of any missing values. Whereas MI consists
  of imputation, analyses and pooling steps, FIML analyses the data in a
  single step. When the assumptions are met the two approaches should
  produce equivalent results. {[}REF{]} As FIML requires specialised
  software, not all analyses can be performed with standard software.
  {[}REF{]}
\item
  Add table describing 2l- imputation models. like this
  \url{https://bookdown.org/mwheymans/bookmi/multiple-imputation-models-for-multilevel-data.html}
\end{itemize}

\hypertarget{imputation-process}{%
\subsubsection{Imputation process}\label{imputation-process}}

\hypertarget{clustering-specification}{%
\paragraph{Clustering specification}\label{clustering-specification}}

\hypertarget{selection-of-variables-in-the-imputation-model}{%
\paragraph{Selection of variables in the imputation
model}\label{selection-of-variables-in-the-imputation-model}}

\begin{itemize}
\tightlist
\item
  Inclusion of outcome variable {[}Moons{]} When the outcome is
  time-to-event, the Nelson-Aalen estimate of the time to event should
  be included as a covariate in the imputation model {[}REF{]}
\item
  Imputation for interaction term, JAV, Machine lerarning methods
  (Random forest) and imputation separated by treatment group
\item
  Prediction matrix specfication (table)
\end{itemize}

\hypertarget{convergency-problems-during-imputation}{%
\paragraph{Convergency problems during
imputation}\label{convergency-problems-during-imputation}}

\begin{itemize}
\tightlist
\item
  Check log
\item
  Problems of convergence on some 2l methods can be solved by use of
  overflux or quickpred.
\item
  Consider the use of scaled variables Vanburen.
\item
  Consider to swap imputation mehtod in problematic variables: The use
  of non parametrical imputation methods such as pmm. or downgrade to 1l
  methods.
\item
  Also by the specification of post-processing in some variables can
  alivianate this.
\item
  Separated imputation process for each endpoint analysis
\end{itemize}

\hypertarget{convergency}{%
\subsubsection{Convergency}\label{convergency}}

Plots and also the prediction plots.

\hypertarget{sensibility-analysis}{%
\subsubsection{Sensibility analysis}\label{sensibility-analysis}}

About the possibility of assumptions about Missing mechanism.

@@

In this example, we demonstrate a multilevel imputation of random
intercept and random slope model with a continuous response. We utilize
the obesity dataset included in the \texttt{micemd}@ package, a
simulated dataset that emulates an electronic survey in which
individuals are asked to provide information about their weight and
consumption habits in different countries. We simulate data for 5
clusters so that the true values are known. We use the following
variables from the dataset:

\begin{itemize}
\tightlist
\item
  \textbf{Cluster:} Region of the patients' healthcare provider (Cluster
  variable),
\item
  \textbf{Gender:} Subjects' Gender (0=male, 1=female),
\item
  \textbf{Age:} Subjects' age,
\item
  \textbf{Height:} Subjects' height in metres,
\item
  \textbf{Weight:} Subjects' weight in kilograms,
\item
  \textbf{BMI:} Subjects' body mass index,
\item
  \textbf{FamOb:} Family obesity history (yes or no),
\item
  \textbf{Time:} Response time in minutes (exclusion-restriction
  variable).
\end{itemize}

@@

In the primary analysis model, BMI serves as the dependent variable,
with Age, Gender, and FamOb as predictors. Because of the clustered
nature of the data, which is quantified with the Intraclass Correlation
Coefficient (ICC) below, we include random intercepts, as well as a
random slope for the Age variable. The model is represented as:
\begin{equation}
\label{eqn:main}
BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
\end{equation}

We start by loading the data:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(ggpubr)
R> myplots <- lapply(1:5, function(i) {
+   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
+   ggplot2::ggtitle(paste0("Cluster", i))
+  })
R> ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1}

\caption[Missing pattern]{Missing pattern}\label{fig:obesity-md}
\end{figure}
\end{CodeChunk}

We observe that the missing pattern is non-monotonic and quite similar
across the clusters. However, regarding the weight variable, we notice
that is systematically missing in cluster 3. In order to evaluate if we
require a imputation method that accounts for clustering we assess the
Intraclass Correlation

\begin{CodeChunk}
\begin{CodeInput}
R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
R> #Performance::ICC(Nulmodel)
\end{CodeInput}
\end{CodeChunk}

Since the ICC is above 0.1 and as the main analysis will be use a mixed
model, we decide to use two-level (2l) imputation methods. In this
imputation process, we include all predictor variables from equation
\ref{eqn:main} in the main model. However, since BMI is a composite of
weight and height, we use deterministic imputation for these, which is
described below.

We use the \textbf{find.defaultMethodfunction} provided in the
\textbf{micemd} package, which suggests an appropriate method for MAR
variables based on the type of variable, number of observations in the
cluster, and number of clusters.

It suggests using `2l.2stage.bin' for the FAV variable and
`2l.2stage.norm' for the age variable. However, after inspecting the age
density plot, we consider modifying its method to `2l.2stage.pmm'. For
the BMI variable, we employ deterministic imputation.

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+                                ni.small = 100, prop.small = 0.4)
R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
R> meth_mar["Age"]<-"2l.2stage.pmm" 
\end{CodeInput}
\end{CodeChunk}

For these imputation models, it is necessary to specify the prediction
matrix, with the cluster variable labelled as -2 and the predictor
variable measured within clusters labelled as 2, encompassing all
variables. We need to suprime the variable Time as this variable is not
specified in the main model. We also modify the relationship between
BMI, weight and height in the prediction matrix to avoid circular
predictions. Then we proceed to run the imputation model.

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mar <- quickpred(Obesity)   # predictor matrix
R> pred_mar[,"Cluster"] <- -2 # clustering variable
R> pred_mar[,"Time"] <- 0
R> pred_mar[pred_mar==1] <- 2
R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
R> ggmice::plot_pred(pred_mar)
\end{CodeInput}

\begin{center}\includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmar-1} \end{center}

\begin{CodeInput}
R> imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+                 m=10, seed = 123, printFlag = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 -9.826  69.500  82.789  82.337  94.919 148.712 
\end{CodeOutput}
\end{CodeChunk}

We are also contemplating the utilisation of the ???pmm??? option, as
the values imputed using a fully parametric method may be implausibly
low for some patients.

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mar["Weight"]<-"2l.2stage.pmm" 
R> imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+                     m=10, seed = 123, printFlag = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar_pmm,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.35   67.80   80.53   80.90   92.94  134.61 
\end{CodeOutput}
\begin{CodeInput}
R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
\end{CodeInput}
\end{CodeChunk}

After confirming convergence, we proceed to save the results for future
use. We consider the possibility that patients may not have been
selected randomly, which would then have led to a distribution for
weight that does not reflect the weight in the population. It???s likely
that an omitted variable, like self-esteem, could influence this
selection. For instance, individuals with lower self-esteem might have
higher weight values, impacting their willingness to provide honest
information due to embarrassment.

To address this situation, two approaches have been proposed for dealing
with Missing Not at Random (MNAR) data: pattern-mixed models and
selection models. Within pattern-mixed models, methods like the delta
method and more advanced ones like NARFS have been suggested. The
selection model approach includes methods such as the Heckman model,
which can be particularly useful in this case. Several methods,
including those by \cite{Galimar_2017,Hammon_2021}, and the recently a
Heckman method designed for two-level data, allow for variations in
intercepts and exposure effects (random intercept and slope).

To apply the \textbf{2l.2stage.heckman} method, the weight variable
should be specified as `2l.2stage.heckman' found in the micemd package.
Additionally, the prediction matrix needs modification because this
method involves specifying two equations: one for the outcome,
describing the incomplete variable in terms of partially observed
predictors (in this case, all variables from the main model), and the
other for the selection model, explaining the probability of being
observed based (R) on certain variables. For the outcome equation we
consider the same imputation model that we used for the MAR case (main
model).
\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
Regarding the selection equation, we include the same predictors as
those in the main model, as well as a time variable. Here the time
variable serves as a restriction exclusion variable specifically
explaining the probability of being observed but not affecting the
incomplete value (Weight). In this context, we assume that the time a
user spends completing the survey serves as a proxy for the barriers
they may encounter in survey completion, such as familiarity with the
survey content or internet speed. These factors may lead the user to
skip specific questions or even the entire survey. Also, we assume the
time does not have any influence on the subject???s weight.
\[R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]

These two equations are jointly estimated under the assumption that the
error terms are interconnected with a bivariate normal distribution. For
a more comprehensive understanding of the model and the exclusion
restriction, see \cite{Munoz_2022}.

To use information from both equations, we must adjust the prediction
matrix. The cluster variable remains specified as before (-2). In this
imputation method, all the variables present in both the selection and
outcome equations are included with a random effect.

However, it is essential to distinguish which of these variables appear
in each equation. In this framework, when a variable is shared between
both equations, it is denoted as (2). Predictors exclusive to the
outcome equation are indicated as (-4), while those exclusive to the
selection equation are labelled as (-3). Consequently, the only
alteration needed in the predictor matrix pertains to the variable
`Time'.

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mnar <- pred_mar
R> pred_mnar["Weight","Time"]<- -3
R> #ggmice::plot_pred(pred_mnar)
\end{CodeInput}
\end{CodeChunk}

We also need to modify the method of the weight variable.

@@

Then we proceed to run the imputation model as before, after executing
these imputation procedures, it is essential to assess convergence and
the coherence of the imputed values.

\begin{CodeChunk}
\begin{CodeInput}
R> imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+                 m=10, seed = 123, printFlag = FALSE)
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 -7.334  69.553  83.722  83.086  96.738 183.338 
\end{CodeOutput}
\end{CodeChunk}

Upon examining the weight variable, we noticed that the imputed range
falls outside the realm of plausible values (as weight should be
positive).

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 -7.334  69.553  83.722  83.086  96.738 183.338 
\end{CodeOutput}
\end{CodeChunk}

Consequently, as before we use the `pmm', option but this time for the
Heckman imputation, this approach ensures that the imputed values remain
within the range of observable values. We then run the imputation model
but this time using the option of pmm, to assure that weight values are
in the range of the observable data, this can be implemented by setting
the pmm parameter to true.

\begin{CodeChunk}
\begin{CodeInput}
R> imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+                      m=10, seed = 123, pmm = T,  printFlag = FALSE)
\end{CodeInput}
\end{CodeChunk}

We check the convergency of the results

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar_pmm,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.35   71.68   85.07   85.07   98.21  134.61 
\end{CodeOutput}
\begin{CodeInput}
R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
\end{CodeInput}
\end{CodeChunk}

After this modification we proceed to compare the effects on the model.
We run the analysis model on each of the completed datasets as well
asthe dataset where the incomplete values are removed (Complete Case
analysis, CC).
